Management induced changes of soil organic carbon on global croplands

### 2020-11-09 SOCbudgetPaper/SOCbudget.tex
update with spell check, references for discussion part, additional analysis on initialization, update of figures

%% 2-column papers and discussion papers
\documentclass[gc, manuscript]{copernicus}

@@

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% ASTRA Proceedings (ap)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

@@

@@

\correspondence{Kristine\ Karstens\ (kristine.karstens@pik-potsdam.de)}

@@

\begin{abstract}
Soil organic carbon (SOC) is one of larges carbon stocks on earth. It is
three times larger than the biospheric pool and more than twice as big
as the athmospheric pool, when looking into the first meter of the earth
soil profile. Human cropping activties led and still lead to a depletion
of SOC though, which are so far not well represented in global
assessments of historic carbon emissions. While SOC models often
represent well the biochemical processes that lead to the accumulation
and decay of SOC, the management decisions driving these biophysical
processes are still little investigated on the global scale. Here we
create a spatial explicit data set for agricutural management on
cropland, especially for crop residue and manure management, based on
global historic production (FAOSTAT), and land-use (LUH2) data; and
combine it with the IPCC Tier 2 approach to create a half degree
resolution SOC stock changes on mineral soils. We estimate that due to
arable farming soils have lost over 37 GtOC of which 4 GtOC have been
regained within the period 1975-2010. We show that, our results on
global scale based on Tier 2 IPCC methodolgy are in good agreement with
Tier 1 default assumptions. We also find that SOC is very sensitive to
management decision such as residue recycling indicating the nessessity
to incorporated better management data in soil models.
\end{abstract}

@@

Soil Organic Carbon (SOC), the amount of organic carbon stored through
the earth's soil, is the largest terrestrial carbon pool, exceeding the
carbon in the atmospheric and vegetation pools multiple times (Batjes).
As such, even small changes in drivers of SOC may thus lead to
substantial shifts in earth carbon cycle and influence the amount of CO2
in the atmosphere (ref. permafrost melting). The specific amount of
carbon stored in the soil is, however, uncertain, with estimates ranging
from 1500 to 2400 GtC for the first meter of the soil profile (Batjes,
1996; more). The quality of SOC maps has markedly improved in recent
decades, along with better understanding of the factors driving the
global magnitude, distribution, and dynamics of SOC pools. Natural
properties like climatic, biophysical, and landscape characteristics
clearly play the most important roles to determine SOC variations over
space and time. Human intervention, including land cover change and land
management, has however added a further driver to SOC change, which
alters terrestrial carbon pools in much shorter time scales and is
likely one of the most dominant driver of SOC changes on managed land
today (cite). Recent studies identify the anthropogenic SOC debt for the
first meter of the soil profile at around 116 GtC (Sanderman et al.),
compared to previous estimates between 60-130 GtC (Lal, 2006). Other
studies have focused more closely on spatially disaggregation of SOC
changes via advanced digital soil mapping techniques (S-World;
Stoorvogel 2, 2017) or better representation of biogeochemical processes
within SOC dynamics (Hararuk, 20XX). While providing a rough estimate of
the order of magnitude of change, these studies lack a detailed
consideration of land management, especially of agricultural activities.
Field-scale models (cite Daycent, RothC, Ecosse, C-Tool) are able to
capture these land management impacts by using detailed information on
crop yield levels, fertilizer inputs and various other on-farm measures.
However, due to the lack of comprehensive global management data as
input to these models, scaling up to the global extent remains a complex
challenge. Our study combines an spatially explicit estimate of
agricultural management data on the global level with a 3-pool SOC model
parametrized for global croplands. This allows us to estimate SOC stock
change factors, as well as organic carbon flow dynamics within the
agricultural system. We thereby consider change in SOC caused by
historical land cover change as well as of different agricultural
management practices, including residue recycling, manure amendments,
irrigation, and tillage. We provide the first global, spatially explicit
SOC loss maps that considers detailed agricultural management. This
paper will first introduce in the methods section the basic concept of
SOC dynamics as applied in this study. We continue with a detailed
description of the global gridded management data used here, including
crop production levels, residue recycling rates, manure amendments, and
the adoption of irrigation and tillage practices. Lastly, we shortly
refer to the concept of stock change factors as outlined in the Tier 1
approach of the IPCC guidelines. In our results section we focus on the
SOC dynamics of global croplands by (1) analysing the spatially explicit
distribution and depletion of SOC from 1975 to 2010, (2) comparing
impacts of different management effects on SOC emissions and (3)
quantify global agricultural carbon flows and stocks to compare the
importance of various management aspects. Finally, we discuss and
compare our findings -- including their implications for SOC model
development -- and conclude with an outlook on the ability of SOC
management to mitigate climate change and contribute to negative
greenhouse gas emissions. \newpage

\section{Method}

We compile calculations as open-source R packages available at
github.com/pik-piam/mrcommons (management related functions) and
github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which
are both based on the MADRaT package (package citation?), a framework
which aims to improve reproducibility and transparency in data
processing. In the following chapter we outine the most important
relationships and assumptions. Table @ref(append:subsection2mrfunctions)
provides further information on corresponding code within the R
packages.

@@

Following the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}; short Tier 2 steady-state method) , we
estimate soil organic carbon (SOC) stocks and their change over time for
cropland on half-degree resolution from 1975 to 2010. We assume the
current SOC state converges towards a steady state, which itself is
depending on biophysical, climatic and agronomic conditions. Therefore
we conduct the following three steps within each yearly timestep: (1) We
calculate annual land-use (sub-)type-specific steady states and decay
rates for SOC stocks, (2) we account for land conversion by transferring
SOC from and to an other land-use type representad as natural
vegetation, (3) we estimate SOC stocks and changes based on the stocks
of the previous time period, the steady state stocks and the decay rate.

\subsubsection{Steady-state SOC stocks and decay rates}

In a simple first order kinetic approach the steady-state soil organic
carbon stocks \(SOC^{eq}\) are given by \begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
(\#eq:inoutflow)
\end{equation} with \(C^{\textrm{in}}\) being carbon inputs to the soil,
\(k\) denoting the soil organic carbon decay rate; as well as \(i\)
representing grid cell indices and \(t\) years. We use for our
calculations the Tier 2 steady-state method, which assumes three soil
carbon sub-pools (active, slow and passive) and interactions between
them. Annual carbon inflow to each sub-pool and annual decay rates of
each sub-pool are the key components to determining steady-state SOC
stocks.

@@

We account for different carbon input sources depending on the two
land-use types we distinguish: croplands and natural vegetated land as
representative for all other land use (see table
@ref(tab:datasourceinputs)). Carbon sources for cropland are recycled
crop residues, below ground biomass of crops (for both see
@ref(sec:residues)) and recycled manure (see @ref(sec:livstmanure)); for
natural vegetation litterfall \citep{schaphoff_lpjml4_2018} is the only
source of carbon inflow to the soil.

Following the IPCC carbon accounting methodology, carbon inputs are
disaggregated into metabolic and structural components depending on
their lignin and nitrogen content (see \citet{ipcc_2019_2019}). For each
component the sum of all carbon input sources is allocated to the
respective SOC sub-pools via transfer coefficients. This implies that
both the amount of carbon as well as its structural composition
determine the effective inflow. Data sources for all considered carbon
inputs as well as for lignin and nitrogen content can be found in table
@ref(tab:datasourceinputs).

@@

The sub-pool specific decay rates are influenced by climatic conditions,
biophysical and biochemical soil properties as well as management
factors that all vary over time (t) and space (i). Following the Tier 2
steady-state method we consider temperature (temp), water (wat), sand
fraction (sf) and tillage (till) effects to account for spatial
variation of decay rates. Thus \(k_{sub}\) is given by

\begin{equation}
\begin{aligned}
& k_{active,t,i}  & = &~ k_{active}  ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ & \cdot~ sf_{t,i}\\
& k_{slow,t,i}    & = &~ k_{slow}    ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ &\\
& k_{passive,t,i} & = &~ k_{passive} ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ & ~ &
(\#eq:decayrates)
\end{aligned}
\end{equation}

For cropland we distinguish the effect of different tillage (see
@ref(sec:tillage)) and irrigation (see @ref(sec:irrigation)) practices
on decay rates, whereas for natural vegetation, we assume rainfed and
non-tilled conditions. Data sources as well as drivers considered for
each land-use types are shown in table @ref(tab:datasourcedecay). To
account for variations of decay rates within each grid cell due to
different tillage and irrigation regimes, average rates based on area
shares are calculated.

@@

                      & Soil disturbance & tillage & [sec:tillage] \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 (\#tab:datasourcedecay)
 \end{table*}

\subsubsection{SOC transfer between land-use types}

@@

\begin{equation}
SOC_{lut,t*} = SOC_{lut,t-1} - \frac{SOC_{lut,t-1}}{A_{lut,t-1}} \cdot  AR_{lut,t} + \frac{SOC_{!lut,t-1}}{A_{!lut,t-1}} \cdot  AE_{lut,t} \qquad \forall\quad sub, i  
(\#eq:ctransfer)
\end{equation}

with \(A\) being the area, \(AR\) the area reduction and \(AE\) the area
expansion for a given land-use type \(lut\). Note that \(!lut\) denotes
the sum over all other land-use types, which decreases in the specific
time step \(t\). Data sources and methodology on land-use states and
changes are described in @ref(sec:landuse).

\subsubsection{Total SOC stocks and stock changes}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
(\#eq:SOCstate)
\end{equation}

Reformulating this equation, we obtain a massbalance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation (1))}}  \qquad \forall\quad lut, sub, i.
(\#eq:steadystate2budget)
\end{equation}

The global SOC stock for each time step can than be calculated via

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ -- land-use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ -- total SOC stock within cell}}
(\#eq:totalstock)
\end{equation}

According to the IPCC guidelines SOC changes can be calculates as the
difference of two (aufeinander folgenden) years (see Eq. 5.0A in
\citep{ipcc_2019_2019}). This however will also include naturally
occured changes due to climatic variance over time. For our study we
will define the absolute and relative SOC changes in relation to a
potential natural state \(SOC_{pnv}\) under the same climatic conditions
at time \(t\) in grid cell \(i\), that is based on the natural
vegetation SOC calculations as defined above without accounting for land
conversion from cropland at any time. The absolute changes
\(\Delta SOC\) and relative changes \(F^{SCF}\) a thus given by

\begin{equation}
\Delta SOC_{t, i} = SOC_{t, i} - SOC_{pnv, t, i}\quad, \qquad \quad  F^{SCF}_{t, i} = \frac{SOC_{t, i}}{SOC_{pnv, t, i}}.
(\#eq:totalstock)
\end{equation}

Note that the absolution changes \(\Delta SOC\) can be also interpreted
as the SOC gap due to human cropping activits; whereas relative changes
\(F^{SCF}\) can be denoted as stock change factors as defined within the
IPCC guidelines of 2006 \citep{ipcc_2006_2006}. Note that the SOC gap is
to the negated cummulative SOC emission also refered to as SOC debt
(cite Sanderman) of mankind.

\subsubsection{Initialisation of SOC pools}

To initialize all SOC sub-pools we assume that cropped land as well as
natural vegetation are in a steady state for the specific configuration
present within the initialization year 1961. We assume after a spin-up
period of 15 years reliable results start from 1975 and improve over
time, since dependency on the initial conditions will decrease.

@@

\hypertarget{sec:landuse}{%
\subsubsection{Landuse and Landuse Change}\label{sec:landuse}}

Land-use patterns are based on the Land-Use Harmonization 2
\citep{Hurtt2020} data set, which we aggregate from quarter degree to
half degree resolution. We disaggregate the five different cropland
subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17
crop groups, applying the relative shares for each grid cell based on
the country and year specific area harvested shares of FAOSTAT data
\citep{faostat_faostat_2016} (see @ref(append:Tableluh2fao2mag) for more
details on the crop group mapping). Land-use transitions are calculated
as net area differences of the land-use data on half-degree.

@@

Using half-degree yield data from LPJmL \citep{schaphoff_lpjml4_2018} as
well as half-degree cropland patterns (see @ref(sec:landuse)) we compile
crop group specific half-degree production patterns. We calibrate
cellular yields with a country-level calibration factor for each crop
group to meet historical FAOSTAT production
\citep{faostat_faostat_2016}. Note that by using physical cropland areas
we account for multiple crop harvest events as well as for fallows.

@@

Crop residue production and management is based on a revised methodology
of \citep{bodirsky_n2o_2012} and key aspects are explained here given
its central role for soil carbon modelling. Starting from harvested crop
production (\(CP\)) estimates and their respective harvested crop area
(\(CA\)), we estimate above-ground (\(AGR\)) and below-ground (\(BGR\))
residual biomass using crop group (\(cg\)) specific harvest indices
(\(HI\)) and root:shoot ratios (\(RS\)) as follows

\begin{equation}
\begin{aligned}
AGR & = CP \cdot HI_{\textrm{slope}} + CA \cdot HI_{\textrm{intercept}}\qquad & \textrm{and} \\
BGR & = (CP + AGR) \cdot RS \qquad                                            & \forall\quad cg, i, t.
(\#eq:resbiomass)
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a yield
and an area dependend fraction \citep{ipcc_2006_2006}. Note that
deviating from \citep{bodirsky_n2o_2012} we use harvested instead of
physical crop area to account for increased residue biomass due to
multiple cropping and decreased amounts on fallow land. We assume that
all BGR are recycled to the soil, whereas AGR can be burned or harvested
for other purposes such as feeding animals \citep{weindl}, fuel or for
material use.

@@

A fixed share of the AGR is assumed to be burned on field depending on
the per-capita income of the country. Following \citep{smil1999}) we
assume 25\% burn share for low-income countries according to worldbank
definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income
(\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for
all middle-income countries depending on their per-capita income.
Depending on the crop group 80--90\% of the residue carbon burned on the
fields are lost within the combustion process \citep{ipcc_2006_2006}.

@@

To transform dry matter estimates into carbon, we compiled crop-group
and plant part specific carbon to dry matter (c:dm) ratios (see
@ref(append:Tablec2dm)).

>>>>

%% 2-column papers and discussion papers
\RequirePackage[2020/02/02]{latexrelease}
\documentclass[gc, manuscript]{copernicus}

@@

% Advances in Geosciences (adgeo)
% Advances in Radio Science (ars)
% Advances in Science and Research (asr)
% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
% Annales Geophysicae (angeo)
% Archives Animal Breeding (aab)
% ASTRA Proceedings (ap)
% Atmospheric Chemistry and Physics (acp)
% Atmospheric Measurement Techniques (amt)
% Biogeosciences (bg)
% Climate of the Past (cp)
% DEUQUA Special Publications (deuquasp)
% Drinking Water Engineering and Science (dwes)
% Earth Surface Dynamics (esurf)
% Earth System Dynamics (esd)
% Earth System Science Data (essd)
% E&G Quaternary Science Journal (egqsj)
% European Journal of Mineralogy (ejm)
% Fossil Record (fr)
% Geochronology (gchron)
% Geographica Helvetica (gh)
% Geoscience Communication (gc)
% Geoscientific Instrumentation, Methods and Data Systems (gi)
% Geoscientific Model Development (gmd)
% History of Geo- and Space Sciences (hgss)
% Hydrology and Earth System Sciences (hess)
% Journal of Bone and Joint Infection (jbji)
% Journal of Micropalaeontology (jm)
% Journal of Sensors and Sensor Systems (jsss)
% Magnetic Resonance (mr)
% Mechanical Sciences (ms)
% Natural Hazards and Earth System Sciences (nhess)
% Nonlinear Processes in Geophysics (npg)
% Ocean Science (os)
% Primate Biology (pb)
% Proceedings of the International Association of Hydrological Sciences (piahs)
% Scientific Drilling (sd)
% SOIL (soil)
% Solid Earth (se)
% The Cryosphere (tc)
% Weather and Climate Dynamics (wcd)
% Web Ecology (we)
% Wind Energy Science (wes)

@@

% Pandoc citation processing

@@

\correspondence{Kristine\ Karstens\ (\href{mailto:kristine.karstens@pik-potsdam.de}{\nolinkurl{kristine.karstens@pik-potsdam.de}})}

@@

\begin{abstract}
Soil organic carbon (SOC) is one of larges carbon stocks on Earth. It is three times larger than the vegetation pool and more than twice as big as the athmospheric pool, when looking into the first meter of the earth soil profile. Human cropping activties led and still lead to a depletion of SOC though, which are so far not well represented in global assessments of historic carbon emissions. While SOC models often represent well the biochemical processes that lead to the accumulation and decay of SOC, the management decisions driving these biophysical processes are still little investigated on the global scale. Here we create a spatial explicit data set for agricutural management on cropland, especially for crop residue and manure management, based on global historic production (FAOSTAT), and land-use (LUH2) data; and combine it with the IPCC Tier 2 approach to create a half degree resolution SOC stock changes on mineral soils. We estimate that due to arable farming soils have lost over 37 GtOC of which 4 GtOC have been regained within the period 1975-2010. We show that, our results on global scale based on Tier 2 IPCC methodolgy are in good agreement with Tier 1 default assumptions. We also find that SOC is very sensitive to management decision such as residue recycling indicating the nessessity to incorporated better management data in soil models.
\end{abstract}

@@

Soil Organic Carbon (SOC), the amount of organic carbon stored through the earth's soil, is the largest terrestrial carbon pool, exceeding the carbon in the atmospheric and vegetation pools multiple times (Batjes). As such, even small changes in drivers of SOC may thus lead to substantial shifts in earth carbon cycle and influence the amount of CO2 in the atmosphere (ref. permafrost melting). The specific amount of carbon stored in the soil is, however, uncertain, with estimates ranging from 1500 to 2400 GtC for the first meter of the soil profile (Batjes, 1996; more). The quality of SOC maps has markedly improved in recent decades, along with better understanding of the factors driving the global magnitude, distribution, and dynamics of SOC pools. Natural properties like climatic, biophysical, and landscape characteristics clearly play the most important roles to determine SOC variations over space and time.
Human intervention, including land cover change and land management, has however added a further driver to SOC change, which alters terrestrial carbon pools over much shorter time scales and is likely one of the most dominant driver of SOC changes on managed land today (cite). Recent studies identify the anthropogenic SOC debt for the first meter of the soil profile at around 116 GtC (Sanderman et al.), compared to previous estimates between 60-130 GtC (Lal, 2006). Other studies have focused more closely on spatially disaggregation of SOC changes via advanced digital soil mapping techniques (S-World; Stoorvogel 2, 2017) or better representation of biogeochemical processes within SOC dynamics (Hararuk, 20XX). While providing a rough estimate of the order of magnitude of change, these studies lack a detailed consideration of land management, especially of agricultural activities.
Field-scale models (cite Daycent, RothC, Ecosse, C-Tool) are able to capture these land management impacts by using detailed information on crop yield levels, fertilizer inputs and various other on-farm measures. However, due to the lack of comprehensive global management data as input to these models, scaling up to the global extent remains a complex challenge. Our study combines a spatially explicit estimate of agricultural management data on the global level with a 3-pool SOC model parametrized for global croplands. This allows us to estimate SOC stock change factors, as well as organic carbon flow dynamics within the agricultural system. We thereby consider change in SOC caused by historical land cover change as well as of different agricultural management practices, including residue recycling, manure amendments, irrigation, and tillage. We provide the first global, spatially explicit SOC loss maps that considers detailed agricultural management.
This paper will first introduce in the methods section the basic concept of SOC dynamics as applied in this study. Additionally, we shortly refer to the concept of stock change factors as outlined in the Tier 1 approach of the IPCC guidelines. Lastly, we continue with a detailed description of the global gridded management data used here, including crop production levels, residue recycling rates, manure amendments, and the adoption of irrigation and tillage practices. In our results section we focus on the SOC dynamics of global croplands by (1) analysing the spatially explicit distribution and depletion of SOC from 1975 to 2010, (2) comparing impacts of different management effects on SOC emissions and (3) quantify global agricultural carbon flows and stocks to compare the importance of various management aspects. Finally, we discuss and compare our findings -- including their implications for SOC model development -- and conclude with an outlook on the ability of SOC management to mitigate climate change and contribute to negative greenhouse gas emissions.
\newpage

\hypertarget{method}{%
\section{Method}\label{method}}

We compile calculations as open-source R packages available at github.com/pik-piam/mrcommons (management related functions) and github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which are both based on the MADRaT package (package citation?), a framework which aims to improve reproducibility and transparency in data processing.
In the following section we outline the most important relationships and assumptions. Table \ref{append:subsection2mrfunctions} provides further information on corresponding code within the R packages.

@@

Following the tier 2 approach of the refinement of IPCC guidelines vol.~4 (\citet{ipcc_2019_2019}; short Tier 2 steady-state method), we estimate soil organic carbon (SOC) stocks and their change over time for cropland on half-degree resolution from 1975 to 2010. We assume the current SOC state converges towards a steady state, which itself is depending on biophysical, climatic and agronomic conditions. Therefore, we conduct the following three steps within each yearly time-step:
(1) We calculate annual land-use (sub-)type-specific steady states and decay rates for SOC stocks,
(2) we account for land conversion by transferring SOC from and to another land-use type represented as natural vegetation,
(3) we estimate SOC stocks and changes based on the stocks of the previous time period, the steady state stocks and the decay rate.

\hypertarget{steady-state-soc-stocks-and-decay-rates}{%
\subsubsection{Steady-state SOC stocks and decay rates}\label{steady-state-soc-stocks-and-decay-rates}}

In a simple first order kinetic approach the steady-state soil organic carbon stocks \(SOC^{eq}\) are given by
\begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation}
with \(C^{\textrm{in}}\) being carbon inputs to the soil, \(k\) denoting the soil organic carbon decay rate; as well as \(i\) representing grid cell identifier and \(t\) years. We use for our calculations the Tier 2 steady-state method, which assumes three soil carbon sub-pools (active, slow and passive) and interactions between them, following the approach in the Century model (Parton et al, 1988). Annual carbon inflow to each sub-pool and annual decay rates of each sub-pool are the key components to determining steady-state SOC stocks.

@@

We account for different carbon input sources depending on the two land-use types we distinguish: croplands and natural vegetated land as representative for all other land use (see table \ref{tab:datasourceinputs}). Carbon sources for cropland are recycled crop residues, below ground biomass of crops (for both see \ref{sec:residues}) and recycled manure (see \ref{sec:livstmanure}); for natural vegetation litterfall (including fine root turnover) \citep{schaphoff_lpjml4_2018} is the only source of carbon inflow to the soil.

Following the IPCC carbon accounting methodology, carbon inputs are disaggregated into metabolic and structural components depending on their lignin and nitrogen content (see \citet{ipcc_2019_2019}). For each component the sum of all carbon input sources is allocated to the respective SOC sub-pools via transfer coefficients. This implies that both the amount of carbon and its structural composition determine the effective inflow. Data sources for all considered carbon inputs as well as for lignin and nitrogen content can be found in table \ref{tab:datasourceinputs}.

@@

The sub-pool specific decay rates are influenced by climatic conditions, biophysical and biochemical soil properties as well as management factors that all vary over time (t) and space (i). Following the Tier 2 steady-state method we consider temperature (temp), water (wat), sand fraction (sf) and tillage (till) effects to account for spatial variation of decay rates. Thus, \(k_{sub}\) is given by

\begin{equation}
\begin{aligned}
& k_{active,t,i}  & = &~ k_{active}  ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ & \cdot~ sf_{t,i}\\
& k_{slow,t,i}    & = &~ k_{slow}    ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ &\cdot~ till_{t,i} ~ &\\
& k_{passive,t,i} & = &~ k_{passive} ~ &\cdot~ temp_{t,i} ~ &\cdot~ wat_{t,i} ~ & ~ &
\label{eq:decayrates}
\end{aligned}
\end{equation}

For cropland, we distinguish the effect of different tillage (see \ref{sec:tillage}) and irrigation (see \ref{sec:irrigation}) practices on decay rates, whereas for natural vegetation, we assume rainfed and non-tilled conditions. Data sources as well as drivers considered for each land-use types are shown in table \ref{tab:datasourcedecay}. To account for variations of decay rates within each grid cell due to different tillage and irrigation regimes, average rates based on area shares are calculated.

@@

                      & Soil disturbance & tillage & [sec:tillage] \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 \label{tab:datasourcedecay}
 \end{table*}

\hypertarget{soc-transfer-between-land-use-types}{%
\subsubsection{SOC transfer between land-use types}\label{soc-transfer-between-land-use-types}}

@@

\begin{equation}
SOC_{lut,t*} = SOC_{lut,t-1} - \frac{SOC_{lut,t-1}}{A_{lut,t-1}} \cdot  AR_{lut,t} + \frac{SOC_{!lut,t-1}}{A_{!lut,t-1}} \cdot  AE_{lut,t} \qquad \forall\quad sub, i  
\label{eq:ctransfer}
\end{equation}

with \(A\) being the area, \(AR\) the area reduction and \(AE\) the area expansion for a given land-use type \(lut\). Note that \(!lut\) denotes the sum over all other land-use types, which decreases in the specific time step \(t\). Data sources and methodology on land-use states and changes are described in \ref{sec:landuse}.

\hypertarget{total-soc-stocks-and-stock-changes}{%
\subsubsection{Total SOC stocks and stock changes}\label{total-soc-stocks-and-stock-changes}}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
\label{eq:SOCstate}
\end{equation}

Reformulating this equation, we obtain a mass balance equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation (1))}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate2budget}
\end{equation}

The global SOC stock for each time step can then be calculated via

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ -- land-use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ -- total SOC stock within cell}}
\label{eq:totalstock}
\end{equation}

According to the IPCC guidelines SOC changes can be expressed as the difference of two consecutive years (see Eq. 5.0A in \citep{ipcc_2019_2019}). This however, will also include naturally occurring changes due to climatic variance over time. For our study we defined the absolute and relative SOC changes in relation to a potential natural state \(SOC_{pnv}\) under the same climatic conditions at time \(t\) in grid cell \(i\), that is based on the natural vegetation SOC calculations as defined above without accounting for land conversion from cropland at any time. The absolute changes \(\Delta SOC\) and relative changes \(F^{SCF}\) a thus given by

\begin{equation}
\Delta SOC_{t, i} = SOC_{t, i} - SOC_{pnv, t, i}\quad, \qquad \quad  F^{SCF}_{t, i} = \frac{SOC_{t, i}}{SOC_{pnv, t, i}}.
\label{eq:totalstock}
\end{equation}

Note that the absolution changes \(\Delta SOC\) can be also interpreted as the SOC gap due to human cropping activities; whereas relative changes \(F^{SCF}\) can be denoted as stock change factors as defined within the IPCC guidelines of 2006 \citep{ipcc_2006_2006}. Note that the SOC gap is equivalent to the negated cumulative SOC emission (\citep{pugh_simulated_2015}) also referred to as SOC debt (\citep{sanderman_soil_2017}) of human land use.

\hypertarget{initialization-of-soc-pools}{%
\subsubsection{Initialization of SOC pools}\label{initialization-of-soc-pools}}

To initialize all SOC sub-pools we assume that cropped land as well as natural vegetation are in a steady state for the specific configuration present within the initialization year 1961. We assume after a spin-up period of 15 years reliable results start from 1975 and improve over time, since dependency on the initial conditions decreases.

Additionally, to the tier 2 approach of the refinement of IPCC guidelines vol.~4 \citep{ipcc_2019_2019} and the detailed analysis of management data coming with it, SOC changes can be estimated using the IPCC tier 1 approach of IPCC guidelines vol.~4 (\citet{ipcc_2006_2006}, \citet{ipcc_2019_2019}). Here, stocks are calculated via stock change factors (\(F^{SCF}\)) given by the IPCC for the topsoil (0-30 cm) and based on observational data. \(F^{SCF}\) are differentiated by different crop, management and input systems (here summarized under \(m\)) reflecting different dynamics under changed in- and outflows without explicitly tracking these. Moreover, \(F^{SCF}\) vary for different climatic zones (\(c\)) specified by the IPCC (see \ref{append:climatemap}). The actual SOC stocks as thus calculated based on a given reference SOC stock by

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F^{SCF}_{c,m} \qquad\forall\quad t,
\label{eq:tier1}
\end{equation}

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into corresponding climate zones \(c\). For this analysis we use the default \(F^{SCF}\) from the Tier 1 method of \citep{ipcc_2006_2006}, and \citep{ipcc_2019_2019} as a comparison and consistency check for our more detailed budget approach.

@@

\hypertarget{sec:landuse}{%
\subsubsection{Land use and Land-use change}\label{sec:landuse}}

Land-use patterns are based on the Land-Use Harmonization 2 \citep{hurtt_harmonization_2020} data set, which we aggregate from quarter degree to half degree resolution. We disaggregate the five different cropland subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17 crop groups, applying the relative shares for each grid cell based on the country and year specific area harvested shares of FAOSTAT data \citep{faostat_faostat_2016} (see \ref{append:Tableluh2fao2mag} for more details on the crop group mapping). Land-use transitions are calculated as net area differences of the land-use data on half-degree.

@@

Using half-degree yield data from LPJmL \citep{schaphoff_lpjml4_2018} as well as half-degree cropland patterns (see \ref{sec:landuse}) we compile crop group specific half-degree production patterns. We calibrate cellular yields with a country-level calibration factor for each crop group to meet historical FAOSTAT production \citep{faostat_faostat_2016}. Note that by using physical cropland areas we account for multiple crop harvest events as well as for fallows.

@@

Crop residue production and management is based on a revised methodology of \citep{bodirsky_n2o_2012} and key aspects are explained here given its central role for soil carbon modelling. Starting from harvested crop production (\(CP\)) estimates and their respective harvested crop area (\(CA\)), we estimate above-ground (\(AGR\)) and below-ground (\(BGR\)) residual biomass using crop group (\(cg\)) specific harvest index values (\(HI\)) and root:shoot ratios (\(RS\)) as follows

\begin{equation}
\begin{aligned}
AGR & = CP \cdot HI_{\textrm{slope}} + CA \cdot HI_{\textrm{intercept}}\qquad & \textrm{and} \\
BGR & = (CP + AGR) \cdot RS \qquad                                            & \forall\quad cg, i, t.
\label{eq:resbiomass}
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a yield and an area dependent fraction \citep{ipcc_2006_2006}. Note that deviating from \citep{bodirsky_n2o_2012} we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased amounts on fallow land.
We assume that all BGR are recycled to the soil, whereas AGR can be burned or harvested for other purposes such as feeding animals \citep{weindl_livestock_2017}, fuel or for material use.

@@

A fixed share of the AGR is assumed to be burned on field depending on the per-capita income of the country. Following \citep{smil_nitrogen_1999}) we assume 25\% burn share for low-income countries according to World Bank definitions (\(<\,1000\,\tfrac{USD}{yr\,cap}\)), 15\% for high-income (\(>\,10000\,\tfrac{USD}{yr\,cap}\)) and linearly interpolate shares for all middle-income countries depending on their per-capita income. Depending on the crop group 80--90\% of the residue carbon burned on the fields are lost within the combustion process \citep{ipcc_2006_2006}.

@@

To transform dry matter estimates into carbon, we compiled crop-group and plant part specific carbon to dry matter (c:dm) ratios (see \ref{append:Tablec2dm}).
